---
title: "Programming Assignment 2"
author: "Joseph Walker - jpwalker625@gmail.com"
date: "March 7, 2017"
output: 
  html_document: 
    toc: yes
---

```{r Global Options, include = TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r Require Packages}
#Load Required Packages
library(tidyjson)
library(jsonlite)
library(lubridate)
library(tidyverse)
library(lubridate)
library(stringr)
```

###Problem 1

Read in the Mixpanel Web Analytics json files and tidy them.

* events: *"berkeley_event-export.json"*

```{r events data set}
events <- "berkeley_event-export.json" %>% gather_array() %>%
  spread_values(name = jstring("name"),
                distinct_id = jstring("distinct_id"),
                time = jstring("time"),
                sampling_factor = jnumber("sampling_factor"),
                browser = jstring("properties", "$browser"),
                browser_version = jnumber("properties", "$browser_version"),
                city = jstring("properties", "$city"),
                current_url = jstring("properties", "$current_url"),
                initial_referrer = jstring("properties", "initial_referrer"),
                initial_referring_domain = jstring("properties", "$initial_referring_domain"),
                lib_version = jstring("properties", "$lib_version"),
                os = jstring("properties", "$os"),
                referrer = jstring("properties", "$referrer"),
                referring_domain = jstring("properties", "$referring_domain"),
                region = jstring("properties", "$region"),
                screen_height = jnumber("properties", "$screen_height"),
                screen_width = jnumber("properties", "$screen_width"),
                artist = jstring("properties", "Artist"),
                duration = jnumber("properties", "Duration"),
                genre = jstring("properties", "Genre"),
                price = jnumber("properties", "Price"),
                title = jstring("properties", "Title"),
                page_name = jstring("properties", "Page Name"),
                Plan = jstring("properties", "Plan"),
                mp_country_code = jstring("properties", "mp_country_code"),
                mp_lib = jstring("properties", "mp_lib")) %>%
                  enter_object("labels") %>%
                  spread_values(labels = jstring("labels"))
```

* people: *"berkeley_people-export.txt"*
(I converted the json file into a text file in order to read it in.)
```{r people data set}
people <- "berkeley_people-export.txt" %>% 
  gather_array() %>% 
  spread_values(distinct_id = jstring("distinct_id"),
                time = jnumber("time"),
                last_seen = jnumber("last_seen"),
                browser = jstring("properties", "$browser"),
                browser_version = jnumber("properties", "$browser_version"),
                city = jstring("properties", "$city"),
                country_code = jstring("properties", "$country_code"),
                email = jstring("properties", "$email"),
                initial_referrer = jstring("properties", "$initial_referrer"),
                initial_referring_domain = jstring("properties", "$initial_referring_domain"),
                name = jstring("properties", "$name"),
                os = jstring("properties", "$os"),
                region = jstring("properties", "$region"),
                timezone = jstring("properties", "$timezone"),
                favorite_genre = jstring("properties", "Favorite Genre"),
                first_login_date = jstring("properties", "First Login Date"),
                lifetime_song_play_count = jnumber("properties", "Lifetime Song Play Count"),
                lifetime_song_purchase_count = jnumber("properties", "Lifetime Song Purchase Count"),
                Plan = jstring("properties", "Plan")) %>%
  enter_object('properties') %>% 
  enter_object("$transactions") %>%
  gather_array() %>% 
  spread_values(transactions_amount = jnumber("$amount"),
                transactions_time = jstring("$time"))
```

Join the two data sets together on the `distinct_id` variable into a third data set.

```{r join events & people}

events_people <- full_join(events, people, by = "distinct_id")
```

Clean up the column names of `events_people` with a function that converts a string to snake case.

```{r snake_case function}

fix_column_names <- function(x){
  s <- gsub("properties\\.\\$", "", x) # remove the 'properties.$' from beginning of colname
  s <- gsub("\\<properties\\>(.)", "", s) # remove properies(.) jargon where
  s <- gsub("\\.", "_", s) # replace '.' with '_'
  s <- gsub("(.)([A-Z][a-z]+)", "\\1_\\2", s) # separate with underscores on capitalization
  s <- tolower(gsub("([a-z0-9])([A-Z])", "\\1_\\2", s)) # lowercase
  s <- gsub("__", "_", s) # double underscore to single underscore
  s <- gsub("^[_, .]", "", s) # delete first char underscore "_" or period "."
  s <- gsub(' ', '', s) # remove spaces
}


colnames(events_people) <- colnames(events_people) %>% fix_column_names()
```

Replace NA's for `artist` and `page_name` with 'None' and 'No Page Name', respectively.

```{r replace NAs}
events_people$artist[is.na(events_people$artist)] <- 'None'

events_people$page_name[is.na(events_people$page_name)] <- "No Page Name"
```

Plot the number of page views grouped by page_name & artist.

```{r Bar Plot}
p <- ggplot(events_people, aes(x = page_name, fill = artist)) +
  geom_bar() + 
  labs(title = "Number of Page Views Grouped by Page Name & Artist",
       caption = "Mixpanel Web Analytics",
       x = "page name",
       y = "page views")
  
p
```

###Problem 2

Read in the AOL search data and fix the column names with the snake case function.
```{r aol_data}

aol_data <- read.table("user-ct-test-collection-01.txt", header = TRUE,sep = "\t")

colnames(aol_data) <- colnames(aol_data) %>% fix_column_names()
```

Sessionize the search records.

* A user session is all events for a given user where there is no more than 30 minutes between events.

Generate the following session statistics:

* anon_id
* session_sequence_number
* session_id
* number_searches
* session_started_at
* session_length
* number_clicks
* mean_item_rank
* mean_number_search_terms

```{r sessionise}
aol_data$query_time <- parse_date_time(aol_data$query_time, "%Y-%m-%d %H:%M:%S")
aol_data$query <- as.character(aol_data$query)
aol_data$query_count <- str_count(strsplit(aol_data$query, " "),pattern = "\\S+")


df <- aol_data %>%
  group_by(anon_id) %>%
  arrange(query_time) %>%
  mutate(minutes_since_last_event = difftime(time1 = query_time, 
                                             time2 = lag(query_time),
                                             units = 'mins'),
         event_boundary = is.na(minutes_since_last_event) | minutes_since_last_event > 30,
         session_sequence_number = cumsum(event_boundary),
         session_id = paste(anon_id, session_sequence_number, sep = "_")) %>%
  group_by(anon_id, session_sequence_number, session_id,query) %>%
  summarise(number_searches = n(),
            session_started_at = first(query_time),
            session_ended_at = last(query_time),
            session_length = as.numeric(difftime(last(query_time), 
                                                        first(query_time),
                                                        units = "mins")),
            number_clicks = sum(!is.na(click_url)),
            mean_item_rank = mean(item_rank),
            mean_number_search_terms = mean(query_count)) %>%
  ungroup()
```

Plot the statistics by session:

1. The distribution of session durations (histogram count)

```{r ggplot 1}
p1 <- ggplot(df, aes(as.integer(session_length))) +
  geom_histogram(binwidth = 1,
                 fill = 'steelblue',
                 color = 'snow') + 
  scale_y_log10() +
  ggtitle(label = 'Distribution of Session Durations') +
  scale_x_continuous(name = 'Session Duration (in minutes)')
  
p1
```

2. The distribution of the number of clicks per session (histogram count)

```{r ggplot 2}

#filter out the 'number_clicks' high outlier & plot
p2 <- ggplot(filter(df, number_clicks < max(number_clicks)),
             aes(number_clicks)) +
  geom_histogram(binwidth = 1,
                 fill = 'purple',
                 color = 'snow') + 
  scale_y_log10() +
  ggtitle(label = 'Distribution of the Number of Clicks per Session') +
  scale_x_continuous(name = 'Number of Clicks') +
  theme_dark()

p2
```

Summarise the data to the user level:

```{r summarise visitors}
users_df <- df %>%
  group_by(anon_id) %>%
  summarise(number_of_sessions = n(),
            avg_session_length = mean(session_length)) %>%
  ungroup()

```

Plot statistics by user:

3. The distribution of the number of sessions by user (histogram count)

```{r ggplot 3}

p3 <- ggplot(users_df, aes(number_of_sessions)) +
  geom_histogram(binwidth = 1,
                 colour = 'darkgreen',
                 fill = 'snow') + 
  scale_y_log10() +
  ggtitle(label = 'Distribution of the Number of Sessions per User') +
  scale_x_continuous(name = 'Number of Sessions (per user)')

p3

```

4. The distribution of mean session duration by user (histogram count)

```{r ggplot 4}

p4 <- ggplot(users_df, aes(avg_session_length)) +
  geom_histogram(binwidth = 1,
                 color = 'snow',
                 fill = 'red') +
  scale_y_sqrt() +
  ggtitle(label = 'Distribution of the Average Session Duration by User') +
  scale_x_continuous(name = 'Average Session Duration (by user)')

p4
```
