---
title: "Lecture 3"
author: "Joseph Walker"
date: "2/22/2017"
output: html_document
---
Quiz: dplyr
```{r Quiz Problems, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(dplyr)
library(knitr)
library(ggplot2)
#1
df <- mtcars %>% mutate(hp_rank = cut(hp, breaks = c(1, 100, 150, Inf), labels = c("low", "medium", "high")))

#2
mpg_table <- mtcars %>% 
  mutate(mean_mpg_all = mean(mpg)) %>%
  group_by(cyl) %>%
  mutate(mean_mpg_by_cyl = mean(mpg), above_all = mpg > mean_mpg_all, above_cyl =mpg > mean_mpg_by_cyl) %>%
  summarise(n_above_mean_all = sum(above_all), n_above_cyl = sum(above_cyl))

#3
p <- ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot()
p
p2 <- ggplot(mtcars, aes(x = factor(cyl), y = mpg, fill = factor(gear))) +
  geom_boxplot()
p2

#4
scores <- data_frame(id = c(1:6), 
                     group_id = c(1211, 1311, 2634, 1311, 1211, 2634), 
                     exam_1 = c(67, 77, 45, 92, 83, 70),
                     exam_2 = c(85, 76, 69, 94, 88, 80),
                     exam_3 = c(91, 94, 88, 79, 97, 100))

kable(scores, align = 'c')
```
Let's summarize the data
In the first example we summarize the data into an untidy, wide data table:
```{r summarizing, echo = TRUE}
#mean exam_grade by group id

#untidy way
mean_exam_grades <- scores %>% 
                    group_by(group_id) %>%
                    summarise(exam_1 = mean(exam_1),
                              exam_2 = mean(exam_2),
                              exam_3 = mean(exam_3))

kable(mean_exam_grades, align = 'c')
```
Now, let's look at the tidy version
What makes a data table tidy?

* Each variable is a separate column (score, exam)
* Each observation is a separate row
* Each value is a separate cell

I can do this by using the `gather` function. 
It gathers the rows into distinct columns using a key-value pairing.
```{r tidy_version}
#tidy version
tidy_exam_grades <- scores %>% 
                    gather(key = exam, value = score, exam_1:exam_3) %>%
                    group_by(group_id, exam) %>%
                    summarise(score = mean(score))

kable(tidy_exam_grades, align = 'c', caption = 'Average Exam Score by Group_ID')
```

Say I wanted to convert my original data frame to a long format and back to a wide format. I can use the `spread` function
```{r spread function, echo = TRUE}
long_scores <- scores %>% gather(key = exam, value = score, exam_1:exam_3)
long_scores

wide_score <- long_scores %>% spread(key = exam, value = score)
wide_score
```
But be careful using spread. 
If you don't have an index, (a unique row_id value), you won't be able to spread the data.

```{r spread error, echo = TRUE}

ldf <- long_scores %>% select(-id)
wdf <- ldf %>% spread(key = exam, value = score)
```

