---
title: 'Final Exam Notes'
output: html_notebook
---

## The Tidy Approach to Data Science

Import 
Tidy
Explore
> Transform
  Visualise
  Model
  
Communicate

Four types of data sources:
files (csv, xml, json, xls, geospatial data, images, etc..)
local or remote databases
Web Service via API


Definition of a tidy data set:

* Each variable has its own column
* Each observation must have its own row
* Each value must have its own cell

What is the advantage of tidy data?

* Defines a standard format on which functions (dplyr verbs) may operate
* Lends itself to R's vectorized operations
* This format allows you to focus on asking questions about the data 
rather than fighting to get the data into the right form for different functions.

Two packages for data visualization:
GGPLOT2, 

Aspects of Data Science:
Includes traditional descriptive and inferential statistics
emphasis on goal: explaining causation
predictive analytics
hypothesis testing
statistical learning
Big Data
High Performance Computing Power
Machine Learning
Data Acquisition
Data Exploration (visualization and interactivity via shiny)
Modeling
Reproducible Research

Stages of A Data Science Project:
Define the goal: what problems am I solving?
What information do I need?
Build Model
Evaluate Model
Present Results & Document
Deploy Model
Repeat


## Dplyr for Data Wrangling
Data wrangling includes tidying and transformation
The basic idea here is to analyze data by applying functions representing analytical verbs to tidy data objects, and then feeding the results to other functions

Piping with %>%

```{r}
library(dplyr)
library(okcupiddata)

data(profiles)

glimpse(profiles)

#Filter
females <- profiles %>% filter(sex == "f")

#arrange
females <- females %>% arrange(age)

males <- profiles %>% filter(sex == "m") %>%
  arrange(age, desc(height))

#Slice
slice(males, 1:10)

#return the first and last n() observation/row
slice(males, c(1, n()))


#select
females %>% select(body_type, diet, drinks) %>%
  top_n(10)

females %>% select(starts_with('d'), starts_with('s')) %>%
  slice(1:10)

females %>% select(drinks_alocohol = drinks, ) %>%
  colnames

females %>% rename(career = job)

iris %>% group_by(Species) %>%
mutate_at(vars(Petal.Width, Sepal.Width), funs(min, max))

mtcars %>% select(hp) %>%
   mutate(model = row.names(),
     ranking = cut(hp, c(0, 100, 150, Inf), c("low", "medium", "high")))
```

## Tidyr for gathering, spreading, separating and uniting data

```{r Tidyr}
############Wide vs. Long Format
library(tidyr)

scores <- data.frame(id = 1:6,
                     group_id = c(1311, 2347, 1311, 1311, 2347, 6232), 
                     exam1    = c(67, 62, 74, 60, 70, 73),
                     exam2    = c(77, 77, 71, 81, 73, 72),
                     exam3    = c(75, 67, 66, 75, 80, 76))


widetolong.scores <- gather(scores, key = exam, value = score, ... = c(-1,-2)) 

#be careful using spread... the data frame must have a column which uniquely identifies each row
longtowide.scores <- spread(widetolong.scores, key = exam, value = score)

#separate to separate parts of a column into two 
df <- data.frame(dates  = sample(seq(as.Date('2000/01/01'), as.Date('2017/02/22'), by="day"), 10))

df %>% separate(dates, c('year', 'month', 'day'))

quakes %>% unite(location, c(lat, long), sep = ", ")
```

## JSON - Java Script Object Notation

The tidyjson package takes an alternate approach to structuring JSON data into tidy data.frames. Similar to tidyr, tidyjson builds a grammar for manipulating JSON into a tidy table structure. Tidyjson is based on the following principles:

* Leverage other libraries for efficiently parsing JSON (jsonlite) 
* Integrate with pipelines built on dplyr and the magrittr %>% operator 
* Turn arbitrarily complex and nested JSON into tidy data.frames that can be joined later * Guarantee a deterministic data.frame column structure 
*Naturally handle 'ragged' arrays and / or objects (varying lengths by document) 
* Allow for extraction of data in values or key names Ensure edge cases are handled correctly (especially empty data)

```{r JSON}
library(tidyjson)
library(jsonlite)

moment <- read_json("moment.json")

df <- moment %>% 
  enter_object('days') %>% 
  gather_array() %>% 
  spread_values(minuteCount = jnumber('minuteCount'),
                pickupCount = jnumber('pickupCount'),
                date = jstring('date'),
                sessions = jstring('sessions'),
                pickups = jstring('pickups'),
                appusages = jstring('appUsages'))
  

sessions <- moment %>%
  enter_object('days') %>%
  gather_array() %>%
  spread_values(minuteCount = jnumber('minuteCount'),
                pickupCount = jnumber('pickupCount'),
                date = jstring('date')) %>%
  enter_object('sessions') %>% 
  gather_array %>%
  spread_values(location_accuracy_in_meters = jnumber("locationAccuracyInMeters"),
                                             length_mins = jnumber("lengthInMinutes"),
                                             lat = jnumber("latitude"),
                                             long = jnumber("longitude"),
                                             date_sessions = jstring("date"))

pickups <- moment %>%
  enter_object('days') %>%
  gather_array() %>%
  spread_values(minuteCount = jnumber('minuteCount'),
                pickupCount = jnumber('pickupCount'),
                date = jstring('date')) %>%
  enter_object('pickups') %>% 
  gather_array %>% 
  spread_values(location_accuracy_in_meters = jnumber("locationAccuracyInMeters"),
                long = jnumber("longitude"),
                end_battery_level = jnumber("endingBatteryLevel"),
                length_in_seconds = jnumber("lengthInSeconds"),
                date_pickups = jstring("date"),
                lat = jnumber("latitude"),
                start_battery_level = jnumber("startingBatteryLevel"))


app_usages <-  moment %>%
  enter_object('days') %>%
  gather_array() %>%
  spread_values(minuteCount = jnumber('minuteCount'),
                pickupCount = jnumber('pickupCount'),
                date = jstring('date')) %>%
  enter_object('appUsages') %>%
  gather_array() %>%
  spread_values(app_name = jstring("appName"),
                on_screen = jnumber("onScreen"))

```

## leaflet for mapping
javascript leaflet mapping library

Two advantages in R:
* You can access the power of leaflet, at a fairly high level, without knowing or using javascript
* leaftlet mapping code is R code, so it seamlessly integrates with your R wrangling, visualizaiton, and modeling scripts

```{r leaflet}
library(leaflet)

markers_df <- data.frame(lng = c(-122.260988, -122.268176, -122.2594134), lat = c(37.871824, 37.870111, 37.8707592), colors = c("red", "blue", "purple"))

labs <- c("Campanile@#1 Public University in the US", "U.C. Berkeley Extension Golden Bear Building", "Sather Gate")

m <- leaflet() %>%
  addProviderTiles(provider = providers$Hydda) %>%
  setView(lng = -122.2727, lat = 37.8716, zoom = 12) %>%
  addCircleMarkers(data = markers_df, popup = labs, color = markers_df$colors) %>%
  addLegend(position = "topright", labels = labs, colors = markers_df$colors)
m
```

## Geospatial Data

shapefiles
points, lines, polygons
ESRI - Environmental Systems Research Institute

GEOJson

Geometries
consist of: a type and a collection of coordinates

types:
a point () - single coordinate
LineStrings - minimum of two coordinates
Polygons

rgdal
readOGR(dsn, layer)

spdplyr
library(spdplyr helpful for merging spatial dataframes with non-spatial dataframes)
spmerge

converting sp to Geojson is easy with geojsonio or rmapshaper
geojson_json

rmapshaper
clip counties with ms_clip

lawn
View(lawn_bbox_polygon)

in leaflet... addGeoJSON(geojson = insert_df_here)

## Shiny

ui.R

```{r}
library(shiny)
ui <- shinyUI(fluidPage(
  titlePanel(
    sidebarLayout(
      sidebarPanel(
        sliderInput("bins",
                     "Number of bins:",
                     min = 1,
                     max = 100,
                     value = 30)
      ),
      mainPanel(
        plotOutput("distPlot")
      )
    )
  )
))


```
server.R
```{r server}
server <- shinyServer(function(input,output){
  output$distPlot <- renderPlot({
    x <- faithful[, 2]
    bins <- seq(min(x), max(x), length.out = input$bins + 1)
    
    hist(x, breaks = bins, col = 'steelblue', border = 'darkgray',main = "Histogram of Old Faithful Eruptions")
  })
})

shinyApp(ui = ui, server = server)

shiny::
```


function	widget
actionButton	Action Button
checkboxGroupInput	A group of check boxes
checkboxInput	A single check box
dateInput	A calendar to aid date selection
dateRangeInput	A pair of calendars for selecting a date range
fileInput	A file upload control wizard
helpText	Help text that can be added to an input form
numericInput	A field to enter numbers
radioButtons	A set of radio buttons
selectInput	A box with choices to select from
sliderInput	A slider bar
submitButton	A submit button
textInput	A field to enter text

## reactivity

selectInput("var"
label = "choose variable",
choices = c(insert choices here))
selected = (default selection from choices above)

switch(input$var
choices go here)

reactive({

})

observe({


})
