---
title: "Assignment_3"
author: "Joseph Walker"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: textmate
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
#load packages
library(shiny)
library(tidyverse)
library(forcats)
library(lubridate)
library(rgdal)
library(leaflet)
library(sp)
library(spdplyr)
library(htmltools)

source("fix_column_names.R")
```

###Problem 1
The U.S. Government Department of Labor, Bureau of Labor Statistics (BLS) publishes a Local Area
Unemployent Report, overview available here:

* https://www.bls.gov/lau/lauov.htm

Data format description here:

* https://www.bls.gov/help/hlpforma.htm#LA


Assignment:
Tidy the data set provided and generate four separate ggplot line plots (including points) of the following:

1. Employment
2. Labor Force
3. Unemploment Rate
4. Unemployment


### Tidy the Data
```{r sf_oak_metro data}

#read in data
df<- read_csv("sf_oak_metro.csv", col_names = T)
df
#tidy Data
sf_oak <- gather(df, Date, Value, -`Series ID`)
sf_oak
#uniform column names (lower case)
colnames(sf_oak) <- colnames(sf_oak) %>% tolower()

#rename the factors
sf_oak$`series id` <- sf_oak$`series id` %>% 
                      fct_recode(unemployment_rate = "LAUMT064186000000003",
                                 unemployment = "LAUMT064186000000004",
                                 employment = "LAUMT064186000000005",
                                 labor_force = "LAUMT064186000000006")


#parse date variable with lubridate functions
sf_oak$date <- parse_date_time2(sf_oak$date, '%b! %Y!')
sf_oak$month <- month(sf_oak$date,label = TRUE)
sf_oak$year <- as.factor(year(sf_oak$date))

glimpse(sf_oak)
```

###Plot the Data {.tabset .tabset-fade .tabset-pills}

#### Employment Statistics
```{r employment plot}

sf_oak %>%
  filter(`series id` == 'employment') %>%
  ggplot(aes(month, value, group = year, colour = year)) +
  geom_point() +
  geom_line() +
  ggtitle(label = 'Monthly Employment Statistics for the San Francisco-Oakland-Hayward Metropolitan Area', 
          subtitle = '1990-2016') +
  theme(plot.title = element_text(size = 10, face = 'bold'),
        plot.subtitle = element_text(size = 10))
```

#### Labor Force Statistics

```{r labor force plot}
sf_oak %>%
  filter(`series id` == 'labor_force') %>%
  ggplot(aes(month, value, group = year, colour = year)) +
  geom_point() +
  geom_line() +
  ggtitle(label = 'Monthly Labor Force Statistics for the San Francisco-Oakland-Hayward Metropolitan Area', 
          subtitle = '1990-2016') +
  theme(plot.title = element_text(size = 10, face = 'bold'),
        plot.subtitle = element_text(size = 10))
```

#### Unemployment Statistics

```{r unemployment plot}
sf_oak %>%
  filter(`series id` == 'unemployment') %>%
  ggplot(aes(month, value, group = year, colour = year)) +
  geom_point() +
  geom_line() +
  ggtitle(label = 'Monthly Unemployment Statistics for the San Francisco-Oakland-Hayward Metropolitan Area', 
          subtitle = '1990-2016') +
  theme(plot.title = element_text(size = 10, face = 'bold'),
        plot.subtitle = element_text(size = 10))
```

#### Unemployment Rate

```{r unemployment rate plot}
sf_oak %>%
  filter(`series id` == 'unemployment_rate') %>%
  ggplot(aes(month, value, group = year, colour = year)) +
  geom_point() +
  geom_line() +
  ggtitle(label = 'Monthly Unemployment Rate Statistics for the San Francisco-Oakland-Hayward Metropolitan Area',
  subtitle = '1990-2016') +
  theme(plot.title = element_text(size = 10, face = 'bold'),
  plot.subtitle = element_text(size = 10))

```

### Problem 2 {.tabset .tabset-fade}

The file assignment2_problem2_data_files.zip contains two files:

* A geospatial polygon file of USA counties: `gz_2010_us_050_00_20m.json`

* a tsv file with monthy employment statistics for all California counties for 2016: `california_counties_monthly_employment_2016.tsv`


#### 2.1 - Choropleth Map

Create a leaflet choropleth map showing the unemployment rates (only, with legend) for all California counties for December 2016 in your HTML report, generated from RMarkdown

```{r problem 2.1}
#Read in the geospatial polygon data using the function rgdal::readOGR()
usa.counties.mapping <- readOGR(dsn = "gz_2010_us_050_00_20m.json",
                                dropNULLGeometries = TRUE)

#fix_column_names
colnames(usa.counties.mapping@data) <- fix_column_names(colnames(usa.counties.mapping@data))

#filter data to include only California
cal.counties.mapping <- usa.counties.mapping[usa.counties.mapping@data$state == '06', ]

#spatial polygons data frame for California
glimpse(cal.counties.mapping@data)

employment.stats <- read_tsv("california_counties_monthly_employment_2016.tsv", col_names = T) 

#employment stats data frame
glimpse(employment.stats)

employment.stats$period <- parse_date_time(employment.stats$period, "%Y-%m-%d")
employment.stats$fips_county<- factor(employment.stats$fips_county)

#filter the data to contain only data for December
employment.filtered <- employment.stats %>% filter(month(period) == 12)

cal.merged <- sp::merge(cal.counties.mapping, employment.filtered, by.x = "county", by.y = "fips_county")
```

```{r 2.1 Visualization}
#to determine the bins
pal <- colorBin(reverse = TRUE, palette = "RdYlBu", domain = cal.merged$unemployed_rate, pretty = TRUE)

labels <- sprintf("<strong>%s</strong><br/>%g%%", cal.merged$name, cal.merged$unemployed_rate) %>%
  lapply(htmltools::HTML)


counties.map <- leaflet(data = cal.merged) %>%
  addProviderTiles(provider = providers$OpenStreetMap.France) %>%
  setMaxBounds(lng1 = -125, lat1 = 31, lng2 = -113, lat2 = 43) %>%
        setView(lng = -119.417931, lat = 36.778259, zoom = 5) %>%
  addPolygons(fillColor = ~pal(cal.merged$unemployed_rate),
              weight = 2,
              opacity = 1,
              color = "white",
              dashArray = "3",
              fillOpacity = .7,
              highlight = highlightOptions(weight = 2,
                               fillColor = 'silver',
                               color = 'blue',
                               fillOpacity = 0.7,
                               bringToFront = TRUE),
              label = labels,
              labelOptions = labelOptions(style = list("font-weight" = "normal",
                                                       padding = "3px 8px"),
                                          textsize = "12px",
                                          direction = "auto")) %>%
  addLegend(position = 'topright', 
            pal = pal, 
            values = ~cal.merged$unemployed_rate,
            title = paste('Unemployment Rate (%)', 'For December', sep = '<br/>'),
            opacity = 0.7)

counties.map
```


#### 2.2 - Shiny App {.tabset .tabset-fade}

A Shiny app that allows the user to choose the month (in a dropdown) for which to display the above choropleth map:

* <https://tictacjoe.shinyapps.io/cal_counties_choropleth_leaflet_app/>


```{r shiny app}
usa.counties.mapping <- readOGR(dsn = "gz_2010_us_050_00_20m.json")

colnames(usa.counties.mapping@data) <- fix_column_names(colnames(usa.counties.mapping@data))

cal.counties.mapping <- usa.counties.mapping[usa.counties.mapping@data$state == '06', ]

employment.stats <- read_tsv("california_counties_monthly_employment_2016.tsv", col_names = T) 
colnames(employment.stats) <- fix_column_names((colnames(employment.stats)))

employment.stats$period <- parse_date_time(employment.stats$period, "%Y-%m-%d")

employment.stats$fips_county<- factor(employment.stats$fips_county)

cal.merged <- sp::merge(cal.counties.mapping, 
                        employment.stats, by.x = "county", 
                        by.y = "fips_county",
                        duplicateGeoms = TRUE)

cal.merged$period <- month(cal.merged$period, label = TRUE)

shinyApp(
ui <- navbarPage(title = "California Counties Employment - 2016", id = 'nav',
           tabPanel('Interactive Map',
                    
                    leafletOutput('county.map', 
                                  width = "800px", 
                                  height = "800px"),
                    
                    absolutePanel(id = "controls", 
                                  fixed = TRUE, 
                                  draggable = TRUE, 
                                  top = "auto", 
                                  left = 20, 
                                  right = "auto",
                                  bottom = 100,
                                  width = "auto", 
                                  height = "auto",
                                  selectInput(inputId = "statistic",
                                              label = 'Select Statistic',
                                              choices = c("employed",
                                                          "labor_force",
                                                          "unemployed", 
                                                          "unemployed_rate")
                                              ),
                                  
                                  selectInput(inputId = "months", 
                                              label = "Select Month", 
                                              choices = sort(cal.merged$period) 
                                              ),
                                  
                                  checkboxInput(inputId = "legend",
                                                label = "Show Legend",
                                                value = FALSE)
                                  ) #end of absolute panel
           ), # End of tab panel 1
           
           tabPanel('Data',
                    
                    dataTableOutput('data.table')
                    )# end of tab panel 2
), #end of navbar page

########################################

function(input, output){

  #############################  
  ### Tab - Interactive Map ###

  by.month <- reactive({cal.merged[cal.merged$period == input$months, ]
  })
    
  by.stat <- reactive({
    switch(input$statistic,
             "employed" = by.month()$employed,
             "labor_force" = by.month()$labor_force,
             "unemployed" = by.month()$unemployed,
             "unemployed_rate" = by.month()$unemployed_rate
         )
  })
  
  my_colors <- reactive({
    colorNumeric(palette = "RdYlBu", reverse = TRUE, domain = by.stat())
  })
  
  output$county.map <- renderLeaflet({
    leaflet(data = cal.merged) %>%
      addProviderTiles(provider = providers$OpenStreetMap.France) %>%
      setMaxBounds(lng1 = -125, lat1 = 31, lng2 = -113, lat2 = 43) %>%
      setView(lng = -119.417931, lat = 36.778259, zoom = 6)
  }) #end of output$county.map
  
  observe({
    pal <- my_colors()
    
    add_labels <- paste(by.month()$name,':', by.stat())
    
    leafletProxy('county.map', data = by.month()) %>%
      clearShapes() %>%
      addPolygons(fillColor = ~pal(by.stat()),
                  weight = 2,
                  opacity = 1,
                  color = "white",
                  dashArray = "3",
                  fillOpacity = .7,
                  highlight = highlightOptions(weight = 2,
                                               fillColor = 'silver',
                                               color = 'blue',
                                               fillOpacity = 0.7,
                                               bringToFront = TRUE),
                  label = add_labels,
                  labelOptions = labelOptions(
                    style = list("font-weight" = "bold",
                                 padding = "3px 8px"),
                    textsize = "15px",
                    direction = "auto")
                  )
     
    })
  
  observe({
    proxy <- leafletProxy("county.map", data = by.month())

    proxy %>% clearControls()
    if (input$legend) {
    proxy %>% addLegend(position = "topright",
                        pal = my_colors(),
                        opacity = .7,
                        values = ~by.stat(),
                        title = paste(input$statistic,'for', input$months)
                        )
      }
    }) # End of Observe
  
##################  
### Tab - Data ###
  
  output$data.table <- renderDataTable(data.frame(by.month()[ , c('name', 
                                                                  'period', 
                                                                  input$statistic)]))
    }, #end of function

options = list(height = 1000)

)
```




