---
title: "Assignment_3"
author: "Joseph Walker"
date: "3/19/2017"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###Problem 1
The U.S. Government Department of Labor, Bureau of Labor Statistics (BLS) publishes a Local Area
Unemployent Report, overview available here:
https://www.bls.gov/lau/lauov.htm
Data format description here:
https://www.bls.gov/help/hlpforma.htm#LA
A file named sf_oak_metro.csv in the Canvas Files, Assignment 3 folder has local area unemployment
statistics for the San Francisco-Oakland-Hayward California Metropolitan Statistical Area (MSA) for
the years 1990 - 2016.
Assignment:
From the dataset above, use tidyverse packages to generate an HTML file from RMarkdown with four
separate ggplot line plots (including points) of the following:
1. Employment (numbers)
2. Labor Force (numbers)
3. Unemploment Rate (rate)
4. Unemployment (numbers)
with:
• The above values on the Y axis (may be labeled simply "value")
• Month abbreviations (Jan, Feb, ... Dec) on the X axis
• Showing the values for the years 1990 - 2016 on each plot indicated by color (include a legend
showing color for each year)
• Each plot identified with a title bar or title caption similar to the above wording, for example,
"Employment: San Francisco-Oakland-Hayward CA Metropolitical Statistical Area"
• A brief description of the problem from the above so that a reader can understand your report.


```{r sf_oak_metro data}
#load packages
library(tidyverse)
library(forcats)
library(lubridate)

#read in data
df<- read_csv("sf_oak_metro.csv", col_names = T)


#06 Labor force,
#05 Employment,
#04 Unemployment, and
#03 Unemployment rate.

#tidy Data
sf_oak <- gather(df, Date, Value, 2:325)
sf_oak$`Series ID` <- sf_oak$`Series ID` %>% 
                      fct_recode(unemployment_rate = "LAUMT064186000000003",
                                 unemployment = "LAUMT064186000000004",
                                 employment = "LAUMT064186000000005",
                                 labor_force = "LAUMT064186000000006")
sf_oak$Date <- parse_date_time2(sf_oak$Date, '%b! %Y!')
sf_oak$month <- month(sf_oak$Date)
sf_oak$year <- year(sf_oak$Date)

```

