---
title: "Programming Assignment 1"
author: "Joseph Walker"
date: "February 10, 2017"
output: html_document
---
 
## Programming Assignment 1

Use the tab-delimited gem data set ddset.tsv posted in the Files section under Assignment 1 to do the
following:

* Write an RMarkdown file and use it to generate a knitted HTML file that includes a bar plot of
the count of observations in the data set showing price ranking by cut, where price ranking is
defined as price:
    + up to the 15th percentile: Lowest
    + 16th to the 30th percentile: Moderate
    + 31st to the 75th percentile: Higher
    + above 75th percentile: Expensive

### Hand in three files:
1. Your RMarkdown file: asn1_firstname_lastname.Rmd
2. Knitted HTML with source: asn1_firstname_lastname_with_source.html (with source code, i.e.
chunk option echo = TRUE)
3. Knitted HTML with no source: asn1_firstname_lastname.html (with source code, i.e. chunk
option echo = FALSE)

### Mandatory Specifications
1. Assume ddset.tsv is on the same directory as your .Rmd file. Your HTML file results should be
100% reproducible: I should be able to Knit it to HTML without modification
2. Include in your results the complete problem description text for Programming Assignment 1
above
3. In the YAML section of your .Rmd file, include title "Programming Assignment 1," author
(your name), and date. Of course, output should be HTML
4. Use base R and ggplot2 only in your solution. Use no other packages except ggplot2 (do not
use dplyr!)
5. Follow the style guidelines described by Hadley Wickham in this R style guide
6. Include a short (no more than one paragraph) description of your approach to solving the
problem at the end of your HTML document


```{r Import & Transform, echo=TRUE, message=FALSE, warning=FALSE}
diamonds <- read.csv(file = "ddset.tsv", header = TRUE, sep = "\t" )

quants <- quantile(diamonds$price, probs = c(.15,.30,.75))

diamonds$price_ranking <- cut(diamonds$price,
                              breaks = c(0, quants, max(diamonds$price)),
                              labels = c("Lowest", "Moderate", "Higher", "Expensive"))
```

```{r Visualize, echo=TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
library(pals)
library(ggthemes)

my_pal <- brewer.pastel1(n = 5)

p <- ggplot(diamonds, aes(price_ranking, fill = cut)) +
  geom_bar() +
  scale_fill_manual(values = my_pal) +
  ggtitle(label = "Diamonds: Price Ranking by Cut") +
  ggthemes::theme_economist_white()

p

```

1. The first step in this assignment requires importing the data set `ddset.tsv`. I have named the data set `diamonds`.
2. Next, I use the `quantile` function on the price variable to determine the cut off values for the specified categories we want to create in the next step.
3. I then use the `cut` function to split the price column into categories or bins based on the price groups we are interested in. I assign these values to a new variable: `price_ranking` into our existing data frame `diamonds`.
4. Finally, I use ggplot to create a stacked bar chart displaying the count of each diamond cut based on its price_ranking.